Architectures for workflow integration in science gateways

### 2016-04-19 paper.tex
Updates in section 4

\emph{Integration} measures the development effort required to build
the architecture. It is obtained by counting the total number of
interactions and components on the architectural diagram. It breaks
down to the folloowing 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}),
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g. account creation, software installation, etc) which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for this criterion, which measures the total number of
software pieces to develop.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or
interactions of the software architecture. It is a critical criterion
since it directly impacts the end users of the science gateway.
Robustness is determined as the number of software components and
interactions that are specific to workflow execution, i.e. not reused
by other functions:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc. 
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red on
the architecture diagrams in Figure~\ref{fig:architectures}. The two
metrics assume that complex architectures tend to be prone to
failure. The two metrics are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of
interaction and component to modify when a new element is
added. Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified. Modification of
a component is required when its code needs to be modified or
recompiled (science gateway or workflow service), or when a new piece
of software has to be installed (infrastructure only). Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new workflow engine type in the
  architecture. Workflow engine belong to different types when they
  cannot be invoked using the same interface. Adding a new type of
  workflow engine allows to execute more workflows in the science gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions
  have to be considered as different engine types. 
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture, assuming that the
  engine type and version required to execute this workflow are
  already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g. GPUs, clouds), or to enforce execution policies
  (e.g. constrain data to remain in a particular network domain).
\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture.

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by reviewing the features
that enable particular aspects of scalability. Features are evaluated
using a 3-level metric: 0 means that the feature is very easy to
enable, 1 means that it can be implemented but with some difficulty,
and 2 means that it is not possible to implement the feature without a
nonsensical amount of effort. Four different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Adding engine instance (\texttt{S$_1$}): this feature
  corresponds to the possibility to add more than 1 engine instance in
  the architecture. Workflow engines may require important amounts of
  resources when a lot of workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load to several engines. \texttt{S$_1$}=0 when
  adding a new engine instance is a fully automated process, i.e. the
  workflow engine only has to be started. In this case, elastic
  engines are possible, i.e. some kind of auto-scaling mechanism can
  be implemented to control the number of engine instances in the
  architecture. \texttt{S$_1$}=1 when adding a new engine instance
  requires some form of manual intervention, which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture and \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues available in the architecture.

\emph{Other features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): ability to describe meta-workflows from existing
  workflows.  \todo{More details about why this is important?} \texttt{O$_1$}=0 when the feature is not available, \texttt{O$_1$}=1 otherwise.
\item Fine-grained debugging (\texttt{O$_2$}): availability of
  debugging information about workflow activities. Fine-grained
  information about workflow activity is required to properly
  troubleshoot workflow executions.  \texttt{O$_2$}=0 when such
  information is easy to access, \texttt{O$_2$}=1
  otherwise. \todo{More details about why this is important?}
\end{itemize}
These 2 metrics are summed to obtain a total number of missing other
features in the architecture.  \todo{Maybe we could revise the
  categories to avoid such a vague ``other features''.}

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), but interactions \texttt{b} and \texttt{c$_2$}
(\texttt{R$_2$}=2) are.

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to modify the science gateway as well as interactions
\texttt{b} and \texttt{c$_2$} (\texttt{E$_1$}=3). Updating a workflow
engine version requires modifications in the science gateway
(\texttt{E$_2$}=1).  Inserting a new workflow is done through
interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new infrastructure generates updates in
interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$}=3).

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
since load-balancing mechanisms available in web servers are usually
meant to deal with short web requests rather than long workflow
executions (\texttt{S$_1$}=2). Distributed engines are not
available by default (\texttt{S$_2$}=1). The scheduling of tasks on
the infrastructure is as complex as in any other architecture since
the workflow engine might implement any kind of scheduling policy
(\texttt{S$_3$}=0).

@@

\paragraph{Extensibility} Adding a new type of workflow engine requires
to implement the corresponding workflow service, to modify interaction
\texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction
(\texttt{E$_2$}=0).  New workflows are added in the science gateway or
in the workflow engine through interaction \texttt{a} only
(\texttt{E$_3$}=1). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{E$_4$}=3).

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new
infrastructure requires to update interactions \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=2).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). Such
scalability properties are one of the major interests of the sub-task
architecture. Distributed engines are not supported by default
(\texttt{S$_2$}=1). Task scheduling is slightly more complex than in
the other approaches due to the special role of the task that executes
the workflow engine (\texttt{S$_3$}=1). Indeed, the reliability of
this task is critical since all the sub-tasks in the workflow depend
on it and, depending on the recovery capabilities of the workflow
engine, may need to be resubmitted if the workflow task fails. The
workflow task is also longer than all its sub-tasks, which increases
its chances of failure. In addition, task parameters, for instance
estimated walltime, are more difficult to estimate for the workflow
task than for sub-tasks which may generate issues such as selection of
wrong batch queues on clusters. Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. Say for instance that only 1 computing resource is
available for the science gateway and that the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen.

@@

\paragraph{Extensibility} Adding a new engine type requires to wrap
the engine in the agent and to update interactions \texttt{b} and
\texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
is transparent (\texttt{E$_2$=0}), and integrating a new workflow is
done through interaction \texttt{a} only
(\texttt{E$_3$=1}). Integrating a new infrastructure requires updates
in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$=3}).

@@

\paragraph{Extensibility} Adding a new type of \emph{parent} engine
requires to implement the corresponding service, to implement
interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
to implement interaction \texttt{d} in the science gateway and in the
parent service (\texttt{E$_1$=5}). Adding a new type of \emph{child}
engine only requires to implement the corresponding service, to
develop interactions \texttt{b} and \texttt{c$_2$} in the child
engine, and to implement interaction \texttt{d} in the parent service
(\texttt{E$_1$=4}). We use \texttt{E$_1$=4.5} in
Table~\ref{table:evaluation} to reflect both conditions. Adding a new
version in the parent or child engine only requires modifying this
engine (\texttt{E$_2$=0}). Adding a new
workflow is done through interaction \texttt{a}
(\texttt{E$_3$=1}). Adding a new infrastructure requires to
re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
interaction \texttt{c$_1$} once (\texttt{E$_4$=5}).

@@

\paragraph{Extensibility} Since adding a new type of workflow engine
aims at supporting more workflows, we consider that it only requires
to re-implement interaction \texttt{g} in this architecture. Note,
however, that implementing interaction \texttt{g} can require very
substantial work depending on the complexity of the language used by
the new engine (\texttt{E$_1$=1}). Based on the same logic, adding a
new engine version only requires modifying interaction \texttt{g}
(\texttt{E$_2$=1}).  Adding a new workflow is done through
interactions \texttt{a} and \texttt{g} (\texttt{E$_3$=2}). As in the
service architecture, interfacing with a new infrastructure requires
modifications in the workflow service and in interactions \texttt{b}
and \texttt{c$_2$} (\texttt{E$_4$=3}).

@@

\textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}7
                                     & \cellcolor[HTML]{99EF99}9
                                     & \cellcolor[HTML]{99FF99}7
                                     & \cellcolor[HTML]{99E099}11
                                     & \cellcolor[HTML]{99BB99}16
                                     & \cellcolor[HTML]{99E899}10\\

@@

  \textbf{Total} (specific software pieces)& \cellcolor[HTML]{99F599}2
                                     & \cellcolor[HTML]{99E199}4
                                     & \cellcolor[HTML]{99CE99}6
                                     & \cellcolor[HTML]{99BB99}8
                                     & \cellcolor[HTML]{99E199}4\\

@@

  New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}1
                                     & \cellcolor[HTML]{99BB99}2\\
New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99BB99}4
                                     & \cellcolor[HTML]{99BB99}4
                                     & \cellcolor[HTML]{99FF99}3
                                     & \cellcolor[HTML]{99BB99}4
                                     & \cellcolor[HTML]{99BB99}4
                                     & \cellcolor[HTML]{99BB99}4\\
  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99C899}9
                                     & \cellcolor[HTML]{99C899}9
                                     & \cellcolor[HTML]{99FF99}7
                                     & \cellcolor[HTML]{99E399}8
                                     & \cellcolor[HTML]{99BB99}9.5
                                     & \cellcolor[HTML]{99E399}8\\

@@

New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2

@@

\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}3
                                     & \cellcolor[HTML]{99DD99}2
                                     & \cellcolor[HTML]{99DD99}2
                                     & \cellcolor[HTML]{99DD99}2
                                     & \cellcolor[HTML]{99DD99}2\\

@@

  \textbf{Total} (missing other features) & \cellcolor[HTML]{99DD99}1

@@

\begin{table*}
\centering
\begin{tabular}{cccccc}
  \textbf{Tight}
  & \textbf{Service}
  & \textbf{Sub-task}
  & \textbf{Pool}
  & \textbf{Nested}
  & \textbf{Import} \\
\cellcolor[HTML]{99E599}2.4
  & \cellcolor[HTML]{99E599}2.5
  & \cellcolor[HTML]{99FF99}1.5
  & \cellcolor[HTML]{99EF99}2.1
  & \cellcolor[HTML]{99BB99}4.0
  & \cellcolor[HTML]{99FA99}1.7\\
\end{tabular}
\caption{Overall evaluation. Brighter colors indicate better performance.}
\label{table:overall}
\end{table*}

>>>>

\emph{Integration effort} measures the development required to build
the architecture. It is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}),
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g. account creation, software installation, etc) which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for this criterion, which measures the total number of
software pieces to develop.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. 
Robustness is determined as the number of software components and
interactions that are specific to workflow execution, i.e. not reused
by other functions:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc. 
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}
assume that complex architectures tend to be prone to
failure. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of
interactions and components to modify when a new element is
added. Modification of a component is required when its code needs to
be modified or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions
  have to be considered as different engine types. 
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions that need to be
  invoked to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g. GPUs, clouds), or to enforce execution policies 
  (e.g. constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture.

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when a lot of workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load to several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention, which prevents easy implementations of elastic
  engines. \texttt{S$_1$}=2 when new engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture and \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.

\emph{Other features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): ability to describe
  meta-workflows from existing workflows.  \todo{More details about
    why this is important?} \texttt{O$_1$}=0 when the feature is
  available, \texttt{O$_1$}=1 otherwise.
\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow activities. When
  such information is available, workflow is seen as a white
  box. Fine-grained information about workflow activity is required to
  properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
  such information is easy to access, \texttt{O$_2$}=1
  otherwise. \todo{More details about why this is important?}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing other
features in the architecture.  \todo{Maybe we could revise the
  categories to avoid such a vague ``other features''.}

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), but interactions \texttt{b} and \texttt{c$_2$} are
(\texttt{R$_2$}=2).

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to modify the science gateway as well as interactions
\texttt{b} and \texttt{c$_2$} (\texttt{E$_1$}=3). Updating a workflow
engine version requires modifications in the science gateway
(\texttt{E$_2$}=1).  Inserting a new workflow is done through
interactions \texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$}=3).

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
(\texttt{S$_1$}=2). Distributed engines are not available by default
(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
as complex as in any other architecture since the workflow engine
might implement any kind of scheduling policy (\texttt{S$_3$}=0).

@@

\paragraph{Extensibility} Adding a new type of workflow engine requires
to implement the corresponding workflow service, to modify interaction
\texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction
(\texttt{E$_2$}=0).  New workflows are added in the science gateway or
in the workflow engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{E$_4$}=3).

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
infrastructure requires to update interactions \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=2).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of the sub-task architecture. Distributed engines are
not supported by default (\texttt{S$_2$}=1). Task scheduling is
slightly more complex than in the other approaches due to the special
role of the task that executes the workflow engine
(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
since all the sub-tasks in the workflow depend on it and, depending on
the recovery capabilities of the workflow engine, may need to be
resubmitted if the workflow task fails. The workflow task is also
longer than all its sub-tasks, which increases its chances of
failure. In addition, task parameters, for instance estimated
walltime, are more difficult to estimate for the workflow task than
for sub-tasks which may generate issues such as selection of wrong
batch queues on clusters. Finally, the interdependencies between the
workflow task and its sub-tasks may create deadlocks when there is
contention. Say for instance that only 1 computing resource is
available for the science gateway and that the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen.

@@

\paragraph{Extensibility} Adding a new engine type requires to wrap
the engine in the agent and to update interactions \texttt{b} and
\texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
is transparent (\texttt{E$_2$=0}) \todo{explain why, also for service}, and integrating a new workflow is
done through interactions \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$=2}). Integrating a new infrastructure requires updates
in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$=3}).

@@

\paragraph{Extensibility} Adding a new type of \emph{parent} engine
requires to implement the corresponding service, to implement
interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
to implement interaction \texttt{d} in the science gateway and in the
parent service (\texttt{E$_1$=5}). Adding a new type of \emph{child}
engine only requires to implement the corresponding service, to
develop interactions \texttt{b} and \texttt{c$_2$} in the child
engine, and to implement interaction \texttt{d} in the parent service
(\texttt{E$_1$=4}). We use \texttt{E$_1$=4.5} in
Table~\ref{table:evaluation} to reflect both conditions. Adding a new
version in the parent or child engine only requires modifying this
engine (\texttt{E$_2$=0}). Adding a new
workflow is done through interaction \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$=2}). Adding a new infrastructure requires to
re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
interaction \texttt{c$_1$} once (\texttt{E$_4$=5}).

@@

\paragraph{Extensibility} Since adding a new type of workflow engine
aims at supporting more workflows, we consider that it only requires
to re-implement interaction \texttt{g} in this architecture  (\texttt{E$_1$=1}). Note,
however, that implementing interaction \texttt{g} can require very
substantial work depending on the complexity of the language used by
the new engine. Based on the same logic, adding a
new engine version only requires modifying interaction \texttt{g}
(\texttt{E$_2$=1}).  Adding a new workflow is done through
interactions \texttt{a$_1$}, \texttt{a$_2$} and \texttt{g} (\texttt{E$_3$=3}). As in the
service architecture, interfacing with a new infrastructure requires
modifications in the workflow service and in interactions \texttt{b}
and \texttt{c$_2$} (\texttt{E$_4$=3}).

@@

\textbf{Total} (total software pieces) & \cellcolor[HTML]{99FF99}\textbf{7}
                                     & \cellcolor[HTML]{99EF99}\textbf{9}
                                     & \cellcolor[HTML]{99FF99}\textbf{7}
                                     & \cellcolor[HTML]{99E099}\textbf{11}
                                     & \cellcolor[HTML]{99BB99}\textbf{16}
                                     & \cellcolor[HTML]{99E899}\textbf{10}\\

@@

  \textbf{Total} (specific software pieces)& \cellcolor[HTML]{99F599}\textbf{2}
                                     & \cellcolor[HTML]{99E199}\textbf{4}
                                     & \cellcolor[HTML]{99FF99}\textbf{1}
                                     & \cellcolor[HTML]{99CE99}\textbf{6}
                                     & \cellcolor[HTML]{99BB99}\textbf{8}
                                     & \cellcolor[HTML]{99E199}\textbf{4}\\

@@

  New workflow -- \texttt{E$_3$} & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99BB99}3\\
New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99E899}3
  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99E099}\textbf{9}
                                     & \cellcolor[HTML]{99E099}\textbf{9}
                                     & \cellcolor[HTML]{99FF99}\textbf{7}
                                     & \cellcolor[HTML]{99EF99}\textbf{8}
                                     & \cellcolor[HTML]{99BB99}\textbf{11.5}
                                     & \cellcolor[HTML]{99EF99}\textbf{8}\\

@@

Multiple engine instances -- \texttt{S$_1$}& \cellcolor[HTML]{99BB99}2

@@

\textbf{Total} (scalability issues)  & \cellcolor[HTML]{99BB99}\textbf{3}
                                     & \cellcolor[HTML]{99DD99}\textbf{2}
                                     & \cellcolor[HTML]{99DD99}\textbf{2}
                                     & \cellcolor[HTML]{99FF99}\textbf{1}
                                     & \cellcolor[HTML]{99DD99}\textbf{2}
                                     & \cellcolor[HTML]{99DD99}\textbf{2}\\

@@

  \textbf{Total} (missing other features) & \cellcolor[HTML]{99DD99}\textbf{1}
                                     & \cellcolor[HTML]{99DD99}\textbf{1}
                                     & \cellcolor[HTML]{99BB99}\textbf{2}
                                     & \cellcolor[HTML]{99DD99}\textbf{1}
                                     & \cellcolor[HTML]{99DD99}\textbf{1}
                                     & \cellcolor[HTML]{99FF99}\textbf{0}\\

@@

\begin{table*}
\centering
\begin{tabular}{cccccc}
  \textbf{Tight}
  & \textbf{Service}
  & \textbf{Sub-task}
  & \textbf{Pool}
  & \textbf{Nested}
  & \textbf{Import} \\
\cellcolor[HTML]{99EE99}2.1
  & \cellcolor[HTML]{99EE99}2.1
  & \cellcolor[HTML]{99FE99}1.5
  & \cellcolor[HTML]{99F499}1.9
  & \cellcolor[HTML]{99BB99}4.0
  & \cellcolor[HTML]{99FF99}1.5\\
\end{tabular}
\caption{Overall evaluation. Brighter colors indicate better performance.}
\label{table:overall}
\end{table*}
